---
title: "Circular Tutorial"
author: "Kees Mulder"
date: "May 25, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Circular data 

First, load the `circular` package:

```{r}
library(circular)
library(ggplot2)
```

### Working with circular data

A simple circular dataset is in the package, which has arrival times in an intensive care unit:

```{r}
fisherB1
ggplot(data.frame(th = fisherB1)) + 
  geom_histogram(aes(x = th, y = ..density..)) 
```

However, importantly, in order to work with it, we must tell the package that the data is in hours, otherwise all the analyses will be wrong:

```{r}
IC_times <- circular(fisherB1, units = "hours")
plot(IC_times)
```

To obtain the mean direction, we use the sum of cosines (\(C = \sum_i \text{cos} (\theta_i) \)), sum of sines (\(S = \sum_i \text{sin} (\theta_i) \)) and \(R = \sqrt{C^2 + S^2} \), which is the square root of the sum of squared $C$ and $S$ values. The mean direction is then given by

\begin{align}
  \overline{\theta}  &= 
    \begin{cases}
      tan^{-1} (S/C)                   &  S > 0, C > 0\\ 
      tan^{-1} (S/C) + \pi             &  C < 0\\ 
      tan^{-1} (S/C) + 2\pi            &  S < 0, C > 0\\ 
    \end{cases}
    \text{, (Fisher, 1995)}
\end{align}
\

In `R`, we can get it, but we have to take the modulus of 24 hours to make this interpretable:

```{r}
mean(IC_times)
mean(IC_times) %% 24
```

### Using circular predictors

Sometimes, we just want to predict a linear outcome $x$ with a circular predictor $\theta$. For example, look at this dataset:

```{r}
fisherB18
```

Then, we can just do:

```{r}
fisherB18$theta <- fisherB18$theta * pi / 180
lmfit <- lm(x ~ cos(theta) + sin(theta), data = fisherB18)
lmfit
```



### Fitting a distribution


We can fit a simple distribution on this data:

```{r}
vm_mles <- mle.vonmises(IC_times)
vm_mles
```

And we can see that the fit is fairly well:

```{r}
ggplot(data.frame(th = IC_times * pi / 12)) + 
  geom_histogram(aes(x = th, y = ..density..)) + 
  stat_function(fun = dvonmises, 
                args = list(mu = vm_mles$mu, kappa = vm_mles$kappa),
                col = "tomato")
```





### Circular outcome regression


```{r}

```



